---
title: "Final Project"
format: html
editor: visual
embed-resources: TRUE
---

#Load in the data/packages

```{r}
ai_impact_data <- read.csv("AI_student_impact.csv")
```

```{r}
library(tidyverse)
library(purrr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(broom)
library(gt)
```

# Dataset description:

This dataset includes sixteen columns and 50000 rows. Each row represents an individual student, shown by their student ID number. The columns include information describing student interactions with AI, spanning academic outcomes, behavioral patterns, institutional context, and psychological well-being. Students are categorized by their major, year of study, pre and post-semester GPA, hours studying with and without use of AI, whether or not they have a paid subscription, their perceived dependency on AI, anxiety level during exams, and burnout risk.

Data dictionary: [Download the Data Dictionary](AI_Impact_Students_Data_Dictionary.docx)

## Question 1: Clean Data

Before analyzing the dataset, we need to ensure each variable has an appropriate data type.

Some variables represent: identifiers, categories, or numerical measurements.

Begin by converting variables to the most appropriate data type.

```{r}
# Code for question 1
clean_ai_impact <- ai_impact_data |> 
  mutate(
    Student_ID = as.character(Student_ID),
    Major_Category = as.factor(Major_Category),
    Year_of_Study = as.factor(Year_of_Study),
    Prompt_Engineering_Skill = as.factor(Prompt_Engineering_Skill),
    Paid_Subscription = as.factor(Paid_Subscription),
    Burnout_Risk_Level = as.factor(Burnout_Risk_Level))

```

Before studying student outcomes, let’s first understand how students are using GenAI tools.

## Question 2: Visualize

Visualize the relationship between students' majors (by category) and average weekly use of generative AI (in hours) using ggplot and a relevant graph/chart.

```{r}
# Code for question 2
clean_ai_impact |> 
  group_by(Major_Category) |> 
  summarize(mean_hours = mean(Weekly_GenAI_Hours)) |> 
  ggplot(aes(x = mean_hours,
             y = Major_Category,
             fill = Major_Category)) +
  geom_bar(stat = "identity") +
  labs(x = "Average Weekly Hours",
       y = "",
       subtitle = "Average Hours of AI Use Weekly by Major Category")

```

## Question 3: Measuring the Correlation Between AI Usage and Student Outcomes

Create a character vector called student_outcomes containing the following variables:

"Post_Semester_GPA" "Skill_Retention_Score" "Anxiety_Level_During_Exams" "Traditional_Study_Hours" "Perceived_AI_Dependency"

Calculate the correlation between Weekly_GenAI_Hours and each outcome variable, and store your results in a tibble with nice formatting and titles.

Describe the results in 2-3 sentences.

```{r}
student_outcomes <- c(
  "Post_Semester_GPA",
  "Skill_Retention_Score",
  "Anxiety_Level_During_Exams",
  "Traditional_Study_Hours",
  "Perceived_AI_Dependency"
)

cor_table <- tibble(
  Outcome = student_outcomes,
  Correlation = map_dbl(
    student_outcomes,
    ~ cor(
      clean_ai_impact$Weekly_GenAI_Hours,
      clean_ai_impact[[.x]]
    )
  )
) |> 
  kable(caption = "Correlations Between Weekly GenAI Usage and Student Outcomes")

cor_table
```

The results suggest that higher weekly GenAI usage is most strongly associated with greater perceived AI dependency, with a relatively strong positive correlation of 0.67. Students who reported using AI more frequently also tended to report slightly higher anxiety during exams.

In contrast, AI usage had weak negative correlations with post-semester GPA, skill retention, and traditional study hours. This may suggest that heavier AI use is associated with spending less time using traditional study methods and slightly lower skill retention, although most of these relationships are fairly weak.


## Question 4: Compute Summary Statistics

Write a function called `numeric_summary` that takes a numeric vector and returns a named vector with the mean, median, and standard deviation. Then use `map()` to apply it to the following columns: `Weekly_GenAI_Hours`, `Traditional_Study_Hours`, `Pre_Semester_GPA`, `Post_Semester_GPA`, and `Skill_Retention_Score`.

```{r}
numeric_summary <- function(x) {
  c(
    mean = round(mean(x, na.rm = TRUE), 3),
    median = round(median(x, na.rm = TRUE), 3),
    sd = round(sd(x, na.rm = TRUE), 3)
  )
}
numeric_cols <- clean_ai_impact |>
  select(Weekly_GenAI_Hours, Traditional_Study_Hours, 
         Pre_Semester_GPA, Post_Semester_GPA, 
         Skill_Retention_Score)

map(numeric_cols, numeric_summary)
```



# Significance Testing

Now that we have a good idea of the students in our dataset, and how much they are using generative AI, let's now test its significance on students' respective GPAs by running simple linear regressions

## Question 5: We now want to test the impact that Generative AI use has had on students' GPAs. Let's begin by running a simple regression with post-semester GPA as the response variable, and traditional study hours (no AI use) as our explanatory variable. (No need for any code, just write in the document.)

::: callout-tip
You may want to look at specific LaTeX notation when writing these in your quarto document!
:::

y = post-semester GPA x = traditional study hours

population equation:

$GPA = \beta_0 + \beta_1 (num_hours) + \varepsilon_i$

sample equation:

$\widehat{GPA} = \hat{\beta}_0 + \hat{\beta}_1 (numhours)$

## Question 6: Before we run our regression, let's check that all the L.I.N.E. conditions are met first.

::: callout-note
You will neeed to create some graphs in order to show that these conditions are satisfied
:::

```{r}
# Run the regression model we just defined and save it to the name "GPA_lm"
GPA_lm <- lm(Post_Semester_GPA ~ Traditional_Study_Hours,
             data = ai_impact_data)

```

```{r}
# Check that the linearity condition is satisfied

ai_impact_data |> 
  ggplot(aes(x = Traditional_Study_Hours,
           y = Post_Semester_GPA)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Number of hours studied weekly without AI Use",
       y = "Student GPA")

```

Although there is a large cluster of points due to number of observations in our data, we still see pretty clear positive linear relationship between study hours and student GPA. We can pretty safely say that the linearity condition is satisfied.

::: callout-note
Since every observation in the data represents a different student and the data is cross-sectional (not time series) the independence assumption is considered to be satisfied
:::

```{r}
# Check that the normality condition is met
plot(GPA_lm, which = 2)

```

Examining the QQ Plot we generated, we can see that while there is some deviation from perfect normality (especially near the tails), given our very large sample size and the fact that our response variable (GPA) is bounded, the normailty condition is reasonably satisfied for the purposes of our analysis.

```{r}
# Check for equal variance
GPA_lm |> 
  augment() |> 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "red", lwd = 1.5)
```

Using the fitted residuals plot here, we can see some issues with the equal variance condition. There seems to be a fan shape to the residuals as opposed to the completely random cluster we hoped to get. They are also not fully clustered around 0 which we would expect to see if the condition was met. However, again, given the amount of observations in our data, and the fact that our response variable is bounded, we can still assume that this condition is satisfied so we can go ahead and use our model to approximate our targeted relationship.

## Question 7: get summary of model

```{r}

summary(GPA_lm)

```
The model shows that Traditional_Study_Hours is a statistically significant predictor of Post_Semester_GPA (p < 0.001). For each hour of traditional study per week, GPA is estimated to increase by approximately [coefficient] points, with an intercept of [intercept]. The R^2 value of [value] suggests that traditional study hours alone explains only a small portion of the variation in GPA.



## Question 8: does some variable improve the model, or not

```{r}
GPA_lm2 <- lm(Post_Semester_GPA ~ Traditional_Study_Hours + Weekly_GenAI_Hours,
              data = ai_impact_data)

summary(GPA_lm2)

anova(GPA_lm, GPA_lm2)
```





## Question 9: Create a formatted regression summary table to compare both models.

Your table should:

-   display the regression coefficients for each model
-   include:
    -   intercept estimates
    -   coefficient estimates
    -   p-values
-   clearly separate the results for each model using grouped row labels
-   rename coefficient terms so they are reader-friendly
-   round coefficient estimates to 3 decimal places
-   display very small p-values as \<.001
-   hide unnecessary regression output columns

::: callout-tip
You will want to store both regression models inside a tibble and consider using a list-column to hold the model objects. Use map() together with broom::tidy() to extract regression coefficients from each model, and use unnest() to expand the coefficient results into rows. Use gt() or kableExtra() to create the final formatted table. View example from week 9 class slides for extra help.
:::

In addition to the table, briefly explain in 1-2 sentences what the results suggest about the predictors of post-semester GPA.

```{r}
# Create a formatted regression table
tibble(
  model = c("GPA_lm", "new_GPA_lm"),
  fit = list(GPA_lm, new_GPA_lm)) |> 
  mutate(coefs = map(fit, ~ broom::tidy(.x))) |> 
  select(model, coefs) |> 
  unnest(cols = coefs) |> 
  mutate(term = fct_recode(
    .f = term,
    "Intercept" = "(Intercept)",
    "Traditional Study Hours" = "traditional_study_hours",
    "Weekly GenAI Usage" = "weekly_genai_usage"
  )) |> 
  select(-std.error, -statistic) |> 
  mutate(p.value = case_when(
    p.value < .001 ~ "<.001",
    TRUE ~ as.character(round(p.value, 3))
  )) |>
  gt(groupname_col = "model") |>
  fmt_number(
    estimate,
    decimals = 3
  ) |>
  tab_row_group(
    label = md("**Model 2: Traditional Study Hours + Weekly GenAI Usage**"),
    rows = model == "Model 2"
  ) |>
  tab_row_group(
    label = md("**Model 1: Traditional Study Hours Only**"),
    rows = model == "Model 1"
  ) |>
  cols_hide(model) |>
  cols_align(
    align = "left",
    columns = term
  ) |>
  tab_style(
    style = cell_fill(color = "gray85"),
    locations = cells_row_groups()
  ) |>
  cols_label(
    term = md("**Model & Term**"),
    estimate = md("**Est. Coef.**"),
    p.value = md("**p-value**")
  )
```

