---
title: "Final Project"
format: html
editor: visual
embed-resources: TRUE
---


#Load in the data/packages

```{r}
ai_impact_data <- read.csv("AI_student_impact.csv")
```

```{r}
library(tidyverse)
library(purrr)
library(knitr)
library(kableExtra)
library(ggplot2)

```

# Dataset description:

This dataset includes sixteen columns and 50000 rows. Each row represents an individual student, shown by their student ID number. The columns include information describing student interactions with AI, spanning academic outcomes, behavioral patterns, institutional context, and psychological well-being. Students are categorized by their major, year of study, pre and post-semester GPA, hours studying with and without use of AI, whether or not they have a paid subscription, their perceived dependency on AI, anxiety level during exams, and burnout risk.

Data dictionary: [Download the Data Dictionary](AI_Impact_Students_Data_Dictionary.docx)

## Question 1: Clean Data

Before analyzing the dataset, we need to ensure each variable has an appropriate data type.

Some variables represent: identifiers, categories, or numerical measurements.

Begin by converting variables to the most appropriate data type.

```{r}
# Code for question 1
clean_ai_impact <- ai_impact_data |> 
  mutate(
    Student_ID = as.character(Student_ID),
    Major_Category = as.factor(Major_Category),
    Year_of_Study = as.factor(Year_of_Study),
    Prompt_Engineering_Skill = as.factor(Prompt_Engineering_Skill),
    Paid_Subscription = as.factor(Paid_Subscription),
    Burnout_Risk_Level = as.factor(Burnout_Risk_Level))

```

Before studying student outcomes, let’s first understand how students are using GenAI tools.

## Question 2: Visualize 
Visualize the relationship between students' majors (by category) and average weekly use of generative AI (in hours) using ggplot and a relevant graph/chart.


```{r}
# Code for question 2
clean_ai_impact |> 
  group_by(Major_Category) |> 
  summarize(mean_hours = mean(Weekly_GenAI_Hours)) |> 
  ggplot(aes(x = mean_hours,
             y = Major_Category,
             fill = Major_Category)) +
  geom_bar(stat = "identity") +
  labs(x = "Average Weekly Hours",
       y = "",
       subtitle = "Average Hours of AI Use Weekly by Major Category")

```

## Question 3: Measuring the Correlation Between AI Usage and Student Outcomes

Create a character vector called student_outcomes containing the following variables:

"Post_Semester_GPA"
"Skill_Retention_Score"
"Anxiety_Level_During_Exams"
"Traditional_Study_Hours"
"Perceived_AI_Dependency"

Calculate the correlation between Weekly_GenAI_Hours and each outcome variable, and store your results in a tibble with nice formatting and titles.

Describe the results in 2-3 sentences.
```{r}
student_outcomes <- c(
  "Post_Semester_GPA",
  "Skill_Retention_Score",
  "Anxiety_Level_During_Exams",
  "Traditional_Study_Hours",
  "Perceived_AI_Dependency"
)

cor_table <- tibble(
  Outcome = student_outcomes,
  Correlation = map_dbl(
    student_outcomes,
    ~ cor(
      clean_ai_impact$Weekly_GenAI_Hours,
      clean_ai_impact[[.x]]
    )
  )
) |> 
  kable(caption = "Correlations Between Weekly GenAI Usage and Student Outcomes")

cor_table
```
The results suggest that higher weekly GenAI usage is most strongly associated with greater perceived AI dependency, with a relatively strong positive correlation of 0.67. Students who reported using AI more frequently also tended to report slightly higher anxiety during exams.

In contrast, AI usage had weak negative correlations with post-semester GPA, skill retention, and traditional study hours. This may suggest that heavier AI use is associated with spending less time using traditional study methods and slightly lower skill retention, although most of these relationships are fairly weak.
