---
title: "Final Project"
author: "Natalie Storz, Jack Fredota, Wendy Muratas"
format: html
editor: visual
embed-resources: TRUE
theme: minty
---

## Load in the data/packages

```{r}
ai_impact_data <- read.csv("AI_student_impact.csv")
```

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(broom)
library(gt)
```

```{r}
ai_impact_data <- read.csv("AI_student_impact.csv")
```

# Dataset description:

The AI Impact on Students dataset contains information on student AI usage habits and academic outcomes, with each row representing a unique student identified by Student_ID. It includes demographic and academic variables such as Major_Category, Year_of_Study, and GPA both before (Pre_Semester_GPA) and after (Post_Semester_GPA) the semester. Several variables capture AI usage behavior, including Weekly_GenAI_Hours, Primary_Use_Case, Prompt_Engineering_Skill, and Tool_Diversity, while Institutional_Policy reflects the student's school-level context. Finally, the dataset includes student wellbeing and outcome measures such as Anxiety_Level_During_Exams, Skill_Retention_Score, Perceived_AI_Dependency, and Burnout_Risk_Level, making it well-suited for exploring how generative AI usage relates to student performance and wellbeing.This dataset includes sixteen columns and 50000 rows.

Data dictionary: <iframe src="AI_Impact_Students_Data_Dictionary.pdf" width="100%" height="600px"></iframe>

## Question 1: Clean Data

Before analyzing the dataset, we need to ensure each variable has an appropriate data type.

Some variables represent: identifiers, categories, or numerical measurements.

Begin by converting variables to the most appropriate data type.

```{r}
# Code for question 1
clean_ai_impact <- ai_impact_data |> 
  mutate(
    Student_ID = as.character(Student_ID),
    Major_Category = as.factor(Major_Category),
    Year_of_Study = as.factor(Year_of_Study),
    Prompt_Engineering_Skill = as.factor(Prompt_Engineering_Skill),
    Paid_Subscription = as.factor(Paid_Subscription),
    Burnout_Risk_Level = as.factor(Burnout_Risk_Level))

```

Before studying student outcomes, let’s first understand how students are using GenAI tools.

## Question 2: Visualize

Visualize the relationship between students' majors (by category) and average weekly use of generative AI (in hours) using ggplot and a relevant graph/chart.

```{r}
# Code for question 2
clean_ai_impact |> 
  group_by(Major_Category) |> 
  summarize(mean_hours = mean(Weekly_GenAI_Hours)) |> 
  ggplot(aes(x = mean_hours,
             y = Major_Category,
             fill = Major_Category)) +
  geom_bar(stat = "identity") +
  labs(x = "Average Weekly Hours",
       y = "",
       subtitle = "Average Hours of AI Use Weekly by Major Category")

```

## Question 3: Measuring the Correlation Between AI Usage and Student Outcomes

Create a character vector called student_outcomes containing the following variables:

"Post_Semester_GPA" "Skill_Retention_Score" "Anxiety_Level_During_Exams" "Traditional_Study_Hours" "Perceived_AI_Dependency"

Calculate the correlation between Weekly_GenAI_Hours and each outcome variable, and store your results in a tibble with nice formatting and titles.

Describe the results in 2-3 sentences.

```{r}
student_outcomes <- c(
  "Post_Semester_GPA",
  "Skill_Retention_Score",
  "Anxiety_Level_During_Exams",
  "Traditional_Study_Hours",
  "Perceived_AI_Dependency"
)

cor_table <- tibble(
  Outcome = student_outcomes,
  Correlation = map_dbl(
    student_outcomes,
    ~ cor(
      clean_ai_impact$Weekly_GenAI_Hours,
      clean_ai_impact[[.x]]
    )
  )
) |> 
  kable(caption = "Correlations Between Weekly GenAI Usage and Student Outcomes")

cor_table
```

The results suggest that higher weekly GenAI usage is most strongly associated with greater perceived AI dependency, with a relatively strong positive correlation of 0.67. Students who reported using AI more frequently also tended to report slightly higher anxiety during exams.

In contrast, AI usage had weak negative correlations with post-semester GPA, skill retention, and traditional study hours. This may suggest that heavier AI use is associated with spending less time using traditional study methods and slightly lower skill retention, although most of these relationships are fairly weak.

## Question 4: Use iteration to compute summary statistics across multiple numeric columns.

Write a function called `numeric_summary` that takes a numeric vector and returns a named vector with the mean, median, and standard deviation. Then use `map()` to apply it to the following columns: `Weekly_GenAI_Hours`, `Traditional_Study_Hours`, `Pre_Semester_GPA`, `Post_Semester_GPA`, and `Skill_Retention_Score`.

```{r}
numeric_summary <- function(x) {
  c(
    mean = round(mean(x, na.rm = TRUE), 3),
    median = round(median(x, na.rm = TRUE), 3),
    sd = round(sd(x, na.rm = TRUE), 3)
  )
}
numeric_cols <- clean_ai_impact |>
  select(Weekly_GenAI_Hours, Traditional_Study_Hours, 
         Pre_Semester_GPA, Post_Semester_GPA, 
         Skill_Retention_Score)

map(numeric_cols, numeric_summary)
```

On average, students spend about \[value\] hours per week using GenAI and \[value\] hours studying traditionally. Pre and post semester GPAs were similar, and skill retention scores averaged around \[value\].

# Significance Testing

Now that we have a good idea of the students in our dataset, and how much they are using generative AI, let's now test its significance on students' respective GPAs by running simple linear regressions

## Question 5: We now want to test the impact that Generative AI use has had on students' GPAs. Let's begin by using LaTeX do write out our population and sample regression equations with skill retention score (recorded post-semester and scored 1-100) as the response variable, and traditional study hours (no AI use) as our explanatory variable. (No need for any code, just write in the document.)

::: callout-tip
You may want to look at specific LaTeX notation when writing these in your quarto document!
:::

y = skill retention score x = traditional study hours

population equation:

$GPA = \beta_0 + \beta_1 (num_hours) + \varepsilon_i$

sample equation:

$\widehat{GPA} = \hat{\beta}_0 + \hat{\beta}_1 (numhours)$

## Question 6: Since our data is so large, we will need to first subset the data and narrow down our analysis. Filter the data and resave it to "clean_ai_impact" so that it only keeps business majors who reported between 6 and 35 hours of traditional study hours. Additionally, remove all students who reported a Skill Retention Score of 100. Now let's fit our model we just specified and save it to the name "retention_lm", so we can check that all the L.I.N.E. conditions are met.

:::::: callout-note
You will need to create some graphs (not base R) in order to show that these conditions are satisfied

```{r}
# Run the regression model we just defined and save it to the name "retention_lm"
retention_lm <- lm(Skill_Retention_Score ~ Traditional_Study_Hours,
             data = clean_ai_impact)

```

```{r}
# Check that the linearity condition is satisfied

clean_ai_impact |> 
  ggplot(aes(x = Traditional_Study_Hours,
           y = Skill_Retention_Score)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "Number of hours studied weekly without AI Use",
       y = "Student GPA")

```

Although there is a large cluster of points due to number of observations in our data, we still see pretty clear positive linear relationship between study hours and student skill retention score. We can pretty safely say that the linearity condition is satisfied.

::: callout-note
Since every observation in the data represents a different student and the data is cross-sectional (not time series) the independence assumption is considered to be satisfied
:::

```{r}
# Check that the normality condition is met
plot(retention_lm, which = 2)

```

Examining the residual distribution and the overlaying density curve we generated, we can comfortably say this condition is satisfied as well. Due to the distribution being unimodal, centered around 0, and showing no serious strong signs of skewness, the condition will be passed.

```{r}
# Check for equal variance
retention_lm |> 
  augment() |> 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "red", lwd = 1.5)
```

Using the fitted residuals plot here, we can see some issues with the equal variance condition. There seems to be a wedge shape to the residuals as we get closer and closer to scores of 100 due to it being the cap. However, given the fact that the residuals are mostly centered around 0 and show not clear and obvious curves or patterns, we can move forward cautiously and say that the condition is satisfied in order to continue our analysis and try to find some evidence of an association between our variables.

## Question 7: Use the summary() function to produce a summary of the model from Question 6, including the coefficients, standard errors, p-values, and other relevant model statistics. Include a short conclusion about the output.

```{r}

summary(retention_lm)

```

The model shows that Traditional_Study_Hours is a statistically significant predictor of Post_Semester_GPA (p \< 0.001). For each hour of traditional study per week, GPA is estimated to increase by approximately 0.013 points, with an intercept of 3.2. The R\^2 value of .01895 suggests that traditional study hours alone explains only a small portion of the variation in GPA.

## Question 8: Does including weekly AI usage as a predictor improve the model? Create a new model including the variable Weekly_GenAI_Hours and compare it to the model from question 6 using a Type 1 anova test.

::: callout-note
You will need this for an interpretation later in the lab.
:::

```{r}
retention_lm2 <- lm(Skill_Retention_Score ~ Traditional_Study_Hours + Weekly_GenAI_Hours,
                    data = clean_ai_impact)

summary(retention_lm2)

anova(retention_lm, retention_lm2)
```

When Weekly_GenAI_Hour is added to the model, it is a statistically significant predictor of Skill_Retnetion_Score (p \< .001), with each additional hour of weekly GenAI use associated with a decrease of approximately 0.156 points in skill retention. The ANOVA comparison confirms that adding  Weekly_GenAI_Hour significantly improves model fit (F = 476.25, p \< .001), suggesting that weekly AI usage does explain additional variation in skill retention beyond traditional study hours alone. However, the R2 only increases from 0.022 to 0.031, meaning both models still explain very little of the overall variation in skill retention scores.

**\*Referenced Source 5**\*

## Question 9: Create a formatted regression summary table to compare both models.

Your table should:

-   display the regression coefficients for each model
-   include:
    -   intercept estimates
    -   coefficient estimates
    -   p-values
-   clearly separate the results for each model using grouped row labels
-   rename coefficient terms so they are reader-friendly
-   round coefficient estimates to 3 decimal places
-   display very small p-values as \<.001
-   hide unnecessary regression output columns

::: callout-tip
You will want to store both regression models inside a tibble and consider using a list-column to hold the model objects. Use map() together with broom::tidy() to extract regression coefficients from each model, and use unnest() to expand the coefficient results into rows. Use gt() or kableExtra() to create the final formatted table. View example from week 9 class slides for extra help.
:::

In addition to the table, briefly explain in 1-2 sentences what the results suggest about the predictors of post-semester GPA. You may also reference the anova table from question 8 to help draw conclusions.

```{r}
# Create a formatted regression table
tibble(
  model = c("retention_lm", "retention_lm2"),
  fit = list(retention_lm, retention_lm2)) |> 
  mutate(coefs = map(fit, ~ broom::tidy(.x))) |> 
  select(model, coefs) |> 
  unnest(cols = coefs) |> 
  mutate(term = fct_recode(
    .f = term,
    "Intercept" = "(Intercept)",
    "Traditional Study Hours" = "Traditional_Study_Hours",
    "Weekly GenAI Hours" = "Weekly_GenAI_Hours"
  )) |>
  select(-std.error, -statistic) |> 
  mutate(p.value = case_when(
    p.value < .001 ~ "<.001",
    TRUE ~ as.character(round(p.value, 3))
  )) |>
  mutate(model = recode(model,       
    "retention_lm"  = "Model 1: Traditional Study Hours Only",
    "retention_lm2" = "Model 2: Traditional Study Hours + Weekly GenAI Usage"
  )) |>
  gt(groupname_col = "model") |>
  fmt_number(
    estimate,
    decimals = 3
  ) |>
  tab_style(
    style = cell_fill(color = "gray85"),
    locations = cells_row_groups()
  ) |>
  cols_label(
    term = md("**Model & Term**"),
    estimate = md("**Est. Coef.**"),
    p.value = md("**p-value**")
  )
```

The results indicate that Traditional Study Hours is a significant predictor of Skill Retention Score, with each additional hour of traditional studying associated with an increase of approximately 0.341 retention points (p \< .001). Weekly GenAI Hours is also a significant predictor (p \< .001), with each additional hour of GenAI use associated with a decrease of approximately 0.156 retention points after controlling for traditional study hours. The ANOVA comparison shows that adding GenAI usage significantly improves model fit (F = 476.25, p \< .001), indicating that GenAI hours explain additional variation in skill retention beyond traditional study time. However, the overall explanatory power of the model remains modest (R² = .031), suggesting that most variation in skill retention is explained by factors not included in the model.

## Sources:

1.  Class Slides

2.  Notes from Professor Schroth who teachers STAT 324 at Cal Poly

3.  [How to Check Linear Regression Assumptions in R](https://www.statology.org/check-linear-regression-assumptions-in-r/)

4.  [How to insert link to a markdown document (Claude](https://claude.ai/share/da4e516e-234a-4f68-bba1-5bd9d8939bda)

5.  [Comparing Regression Models with anova - YaRr](https://bookdown.org/ndphillips/YaRrr/comparing-regression-models-with-anova.html)
::::::
