---
title: "Final Project"
format: html
editor: visual
embed-resources: TRUE
---

#Load in the data/packages

```{r}
ai_impact_data <- read.csv("AI_student_impact.csv")
```

```{r}
library(tidyverse)
library(purrr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(broom)
```

# Dataset description:

This dataset includes sixteen columns and 50000 rows. Each row represents an individual student, shown by their student ID number. The columns include information describing student interactions with AI, spanning academic outcomes, behavioral patterns, institutional context, and psychological well-being. Students are categorized by their major, year of study, pre and post-semester GPA, hours studying with and without use of AI, whether or not they have a paid subscription, their perceived dependency on AI, anxiety level during exams, and burnout risk.

Data dictionary: [Download the Data Dictionary](AI_Impact_Students_Data_Dictionary.docx)

## Question 1: Clean Data

Before analyzing the dataset, we need to ensure each variable has an appropriate data type.

Some variables represent: identifiers, categories, or numerical measurements.

Begin by converting variables to the most appropriate data type.

```{r}
# Code for question 1
clean_ai_impact <- ai_impact_data |> 
  mutate(
    Student_ID = as.character(Student_ID),
    Major_Category = as.factor(Major_Category),
    Year_of_Study = as.factor(Year_of_Study),
    Prompt_Engineering_Skill = as.factor(Prompt_Engineering_Skill),
    Paid_Subscription = as.factor(Paid_Subscription),
    Burnout_Risk_Level = as.factor(Burnout_Risk_Level))

```

Before studying student outcomes, let’s first understand how students are using GenAI tools.

## Question 2: Visualize 

Visualize and compare the distribution of weekly generative AI use across major categories using an appropriate plot

```{r}
# Code for question 2
clean_ai_impact |>
  ggplot(aes(x = Weekly_GenAI_Hours,
             y = reorder(Major_Category, Weekly_GenAI_Hours, median))) +
  geom_boxplot(outlier.shape = NA) +
  labs(x = "Weekly Generative AI Use (Hours)",
       y = "Major Category",
       title = "Weekly Generative AI Use by Major Category")

```

## Question 3: Measuring the Correlation Between AI Usage and Student Outcomes

Create a character vector called student_outcomes containing the following variables:

"Post_Semester_GPA" "Skill_Retention_Score" "Anxiety_Level_During_Exams" "Traditional_Study_Hours" "Perceived_AI_Dependency"

Calculate the correlation between Weekly_GenAI_Hours and each outcome variable, and store your results in a tibble with nice formatting and titles.

Describe the results in 2-3 sentences.

```{r}
student_outcomes <- c(
  "Post_Semester_GPA",
  "Skill_Retention_Score",
  "Anxiety_Level_During_Exams",
  "Traditional_Study_Hours",
  "Perceived_AI_Dependency"
)

cor_table <- tibble(
  Outcome = student_outcomes,
  Correlation = map_dbl(
    student_outcomes,
    ~ cor(
      clean_ai_impact$Weekly_GenAI_Hours,
      clean_ai_impact[[.x]]
    )
  )
) |> 
  kable(caption = "Correlations Between Weekly GenAI Usage and Student Outcomes")

cor_table
```

The results suggest that higher weekly GenAI usage is most strongly associated with greater perceived AI dependency, with a relatively strong positive correlation of 0.67. Students who reported using AI more frequently also tended to report slightly higher anxiety during exams.

In contrast, AI usage had weak negative correlations with post-semester GPA, skill retention, and traditional study hours. This may suggest that heavier AI use is associated with spending less time using traditional study methods and slightly lower skill retention, although most of these relationships are fairly weak.

## Question 4: Compute Summary Statistics

Write a function called `numeric_summary` that takes a numeric vector and returns a named vector with the mean, median, and standard deviation. Then use `map()` to apply it to the following columns: `Weekly_GenAI_Hours`, `Traditional_Study_Hours`, `Pre_Semester_GPA`, `Post_Semester_GPA`, and `Skill_Retention_Score`.

```{r}
numeric_summary <- function(x) {
  c(
    mean = round(mean(x, na.rm = TRUE), 3),
    median = round(median(x, na.rm = TRUE), 3),
    sd = round(sd(x, na.rm = TRUE), 3)
  )
}
numeric_cols <- clean_ai_impact |>
  select(Weekly_GenAI_Hours, Traditional_Study_Hours, 
         Pre_Semester_GPA, Post_Semester_GPA, 
         Skill_Retention_Score)

map(numeric_cols, numeric_summary)
```

# Significance Testing


Now that we have a good idea of the students in our dataset, and how much they are using generative AI, let's now test its significance on students' respective GPAs by running simple linear regressions


## Question 5: We now want to test the impact that Generative AI use has had on students' GPAs. Let's begin by using LaTeX do write out our population and sample regression equations with skill retention score (recorded post-semester and scored 1-100) as the response variable, and traditional study hours (no AI use) as our explanatory variable. (No need for any code, just write in the document.) 

:::{.callout-tip}
You may want to look at specific LaTeX notation when writing these in your quarto document!
:::

y = skill retention score
x = traditional study hours

population equation:

$score = \beta_0 + \beta_1 (numhours) + \varepsilon_i$

sample equation:

$\widehat{score} = \hat{\beta}_0 + \hat{\beta}_1 (numhours)$

## Question 6: Since our data is so large, we will need to first subset the data and narrow down our analysis. Filter the data and resave it to "clean_ai_impact" so that it only keeps business majors who reported between 6 and 35 hours of traditional study hours. Additionally, remove all students who reported a Skill Retention Score of 100. Now let's fit our model we just specified and save it to the name "GPA_lm", so we can check that all the L.I.N.E. conditions are met. 
:::{.callout-note}
You will need to create some graphs (not base R) in order to show that these conditions are satisfied
:::

```{r}
# Run the regression model we just defined and save it to the name "GPA_lm"
clean_ai_impact <- clean_ai_impact |> 
  filter(Traditional_Study_Hours > 5 & Traditional_Study_Hours <= 35,
         Skill_Retention_Score <= 100,
         Major_Category == "Business")
GPA_lm <- lm(Skill_Retention_Score ~ Traditional_Study_Hours,
             data = clean_ai_impact)

```



```{r}
# Check that the linearity condition is satisfied

clean_ai_impact |> 
  ggplot(aes(x = Traditional_Study_Hours,
             y = Skill_Retention_Score)) +
  geom_point() +
  geom_smooth(method = lm) 

```

Although there is a large cluster of points due to number of observations in our data, we still see pretty clear positive linear relationship between study hours and student skill retention score. We can pretty safely say that the linearity condition is satisfied. 

:::{.callout-note}
Since every observation in the data represents a different student and the data is cross-sectional (not time series) the independence assumption is considered to be satisfied
:::

```{r}
# Check that the normality condition is met
resid_sd <- sd(GPA_lm$residuals)

GPA_lm |> 
  broom::augment() |> 
  ggplot(aes(x = .resid)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30) +
  stat_function(fun = ~ dnorm(.x, mean = 0, sd = resid_sd),
                color = "steelblue", lwd = 1.5) +
  xlab("Residuals") +
  theme(aspect.ratio = 1)
```

Examining the residual distribution and the overlaying density curve we generated, we can comfortably say this condition is satisfied as well. Due to the distribution being unimodal, centered around 0, and showing no serious strong signs of skewness, the condition will be passed. 

```{r}
# Check for equal variance
GPA_lm |> 
  augment() |> 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "red", lwd = 1.5)
```

Using the fitted residuals plot here, we can see some issues with the equal variance condition. There seems to be a wedge shape to the residuals as we get closer and closer to scores of 100 due to it being the cap. However, given the fact that the residuals are mostly centered around 0 and show not clear and obvious curves or patterns, we can move forward catiously and say that the condition is satisfied in order to continue our analysis and try to find some evidence of an association between our variables. 

## Question 7: get summary of model


## Question 8: does some variable improve the model, or not


## Question 9: use example from wk 9 notes called "Example Cont. Regression Coefficients" to create a table showing differences in intercepts and coefficients between the 2 models created in Q5 and Q8